A Spiking Independent Accumulator Model for Winner-Take-All Computation

### 2017-04-26 doc/paper/paper.tex
Further paper revisions.

We implement Equation~\ref{eqn:usher-mcclelland} with the NEF by using one population of neurons for each $x_i$, and applying Principle~3 to each population.
By appropriately selecting the gain and bias parameters from Principle~1, we ensure that each state variable is rectified ($x_i \ge 0$) as required.
We believe this implementation is novel as it does not interpret each $x_i$ as a distinct neural firing rate, but rather as a population-level representation distributed across any number of spiking neurons.
In effect, heterogeneous spike trains are weighted by optimal decoding weights to precisely implement the stated dynamics.
%As such, the model equations do not impose restrictions on the neuron model or requires to lump the neurons into a single population firing rate.
This allows us to attain greater biological realism without altering the dynamics prescribed by Equation~\ref{eqn:usher-mcclelland}.
\begin{figure}[t]
    \centering
    \includegraphics{figures/lca}\\
    \vspace*{.3cm}
    \includegraphics{figures/ia}
    \caption{
        Example time course of the state variables in the LCA (top) and IA (bottom) networks with three choices ($D~=~3$).
        The vector of inputs is $\left[ 0.8, 0.7, 0.6 \right]$.
    }\label{fig:usher-mcclelland}\label{fig:indacc}
\end{figure}

\subsection{Independent accumulator model}
The other WTA mechanism that we investigate is our proposed independent accumulator (IA) model (see Fig.~\ref{fig:ia-sketch}).
We use the term `independent' to refer to the fact that there are no direct interactions between each accumulator, unlike in the LCA model which has direct competition between states.
To enable a form of competition, we add a second thresholding layer that  projects back to self-excite and mutually-inhibit the first layer.
We now provide the details of each layer~-- again implemented using the principles of the NEF\@.
\begin{figure}
    \centering
    \includegraphics{figures/ia-sketch}
    \caption{
        Independent accumulator (IA) network.
        Neural populations are denoted by circles labelled with their represented state variable.
        Arrows denote excitatory connections, while lines ending in circles denote inhibitory connections.
        The second layer computes $\bar{x}_i := \Theta(x_i - \vartheta)$.
    }\label{fig:ia-sketch}
\end{figure}

@@

\section{Results}
We find that the ability to form a clear decision of the LCA network decreases with more noise and less target separation  (see Fig.~\ref{fig:decisions}).
Also, the magnitude of the inputs has an important influence.
For a small inputs with $u=0.2$ the winner will mostly not exceed the 0.15 threshold with noise present.
The best performance is achieved with medium inputs with $u=0.6$.
Higher inputs make it more likely that runner-ups will exceed the 0.15 threshold, especially with a small target separation.
In contrast to the LCA network, the IA network manages to form a clear decision in every trial (not explicitly shown in Figure~\ref{fig:decisions}, but all data points fall on the grey line).
\begin{figure*}[t]
    \centering
    \includegraphics{figures/decisions}
    \caption{
        Fraction of trials with a clear decision for the LCA network (which also exactly matches the fraction of correct trials).
        Each plot shows data for a different input magnitude $u$.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal line indicates the optimum, which coincides with the performance of the IA network.
    }\label{fig:decisions}
\end{figure*}

Interestingly, for all clear decisions the correct winner was determined by the LCA network.
Thus, a plot of correct trials looks identical to Figure~\ref{fig:decisions}, with slightly different error bars.
While always reaching a clear decision, the decisions of the IA network are not always correct.
Overall, the IA performance tends to be worse than the LCA performance for a high input magnitude, but better for smaller inputs (Fig.~\ref{fig:correct}a,\,b).
We can greatly improve the IA performance by increasing $\tau_1$ to \SI{0.5}{\second} which slows down the integration of evidence (Fig.~\ref{fig:correct}c).
This improves the IA performance to be above the LCA performance for high baselines, but it will also increase the decisions times of the IA network.
For a low baseline with $u=0.2$, it makes the IA network unable to decide within the allocated time frame, but given more time it would still reach a decision.
%A further improvement to the IA network is possible by eliminating a systematic bias introduced in the second layer of the IA network.
%In the NEF the neuron gain and bias parameters are usually picked from random distributions.
%As such the layer 2 populations will become active at slightly different thresholds.
%By using identical layer 2 populations, i.e., identical gain and bias parameters for each population, this bias can be eliminated and the IA network performs even better (Fig.~\ref{fig:correct}d).
\begin{figure*}[t]
    \centering
    \includegraphics{figures/correct}
    \caption{
        Fraction of correct trials for the IA network.
        Each plot shows data for a different combination of input magnitude $u$ and integration time-constant $\tau_1$.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal line indicates the optimum.
    }\label{fig:correct}
\end{figure*}

As shown in Fig.~\ref{fig:time}a, the time required to reach a decision in the LCA network depends mostly on the size of inputs and target separation.
We averaged over the different noise conditions because the noise influence on the timing was minor.
In the IA network the largest input is the most important factor (dashed vs.\ solid lines in Fig.~\ref{fig:time}b).
Depending on this magnitude, the network can either be faster or slower than the LCA network, but it will need more time given a value of $\tau_1$ that achieves the same fraction of correct responses as the LCA network.
There is also a slight influence of target separation and input noise, with an interaction of these two parameters (solid lines in Fig.~\ref{fig:time}).
\begin{figure*}[t]
    \includegraphics{figures/time1}
    \hfill
    \includegraphics{figures/time2}
    \caption{
        (Left) Mean decision times for the LCA network.
        Shown data is averaged across all noise levels, since noise had minimal effect on decision times.
        (Right) Mean decision times for the IA network with input magnitude $u = 0.2$ (solid) and $u = 1$ (dashed).
        Error bars denote bootstrapped 95\% confidence intervals.
    }\label{fig:time}
\end{figure*}

Finally, looking at the transient responses indicates that both models might produce outputs of losing choices.
For the LCA network the magnitude of the transient response mainly increases with the amount of noise (Fig.~\ref{fig:transient}a).
For the IA network, transient outputs are smaller in noisy conditions, but can be higher than for the LCA network in less noisy conditions with small target separations.
The magnitude of such transient responses is reduced by adjusting $\tau_1$ to 0.5, at the cost of a slower decision.
\begin{figure*}
    \centering
    \includegraphics{figures/transient}
    \caption{
        Transient response (highest output of losing choices) for the LCA and IA network.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal lines show the optimum.
    }\label{fig:transient}
\end{figure*}

@@

Both models might produce a transient response that may be interpreted as a decision, which can make the detection of decisions problematic.
This is of special relevance when incorporating the networks into larger cognitive models.
For the LCA network, the transient behaviour is inherent to its design; there will always be an initial rise of all state variables (that receive input) before the mutual inhibition grows strong enough to push them back to zero.
In the IA network, however, such transient responses may be avoided by choosing appropriate $\tau_1$ and $\vartheta$.
It should be noted that other recurrently connected readout mechanisms with pooled inhibition have been proposed for at least the two choice case~\cite<e.g.>{lo2006} that produce outputs somewhat comparable to the IA network.

@@

We have not yet investigated the agreement of these mechanisms with neurobiological and behavioural data, although this has been done before for other WTA networks~\cite{gold2007,smith2004}.
The implementation in spiking neurons, however, provides some basic biological plausibility and more readily permits comparisons with neural data.
In particular, the IA network predicts that the firing rates for neurons in the first layer will rise up to a threshold, and that neurons in the second layer will not become active or inhibit the first layer until reaching this threshold.
In contrast, the LCA network predicts that the firing of any neurons will proportionately inhibit all other neurons that do not represent the same state.
With regard to behavioural data, different effects for decision times are predicted as the number of choices increases:
for the LCA network decisions become slower due to the mutual interaction, but for the IA network decisions will take less time because only a single accumulator needs to exceed the threshold.
Nevertheless, evidence for one network does not exclude the possibility that the other network is employed for different tasks or by different brain areas.
Relatedly, it might be more plausible to distribute the representation of all state variables over a single population of neurons.
This is directly supported by the NEF, but we chose to leave this to future work to keep the current analysis free from potential interactions of the state variables introduced by such a distributed representation.

>>>>

We implement Equation~\ref{eqn:usher-mcclelland} with the NEF by using one population of neurons for each $x_i$, and applying Principle~3 to each population.
By appropriately selecting the gain and bias parameters from Principle~1, we ensure that each state variable is rectified ($x_i \ge 0$) as required.
We believe this implementation is novel as it does not interpret each $x_i$ as a distinct neural firing rate, but rather as a population-level representation distributed across any number of spiking neurons.
In effect, heterogeneous spike trains are weighted by optimal decoding weights to precisely implement the stated dynamics.
%As such, the model equations do not impose restrictions on the neuron model or requires to lump the neurons into a single population firing rate.
This allows us to attain greater biological realism without altering the dynamics prescribed by Equation~\ref{eqn:usher-mcclelland}.
\begin{figure}[t]
    \centering
    \includegraphics[scale=0.9]{figures/lca}\\
    \vspace*{.3cm}
    \includegraphics[scale=0.9]{figures/ia}
    \caption{
        Example time course of the state variables in the LCA (top) and IA (bottom) networks with three choices ($D~=~3$).
        The vector of inputs is $\left[ 0.8, 0.7, 0.6 \right]$.
    }\label{fig:usher-mcclelland}\label{fig:indacc}
\end{figure}

\subsection{Independent accumulator model}
The other WTA mechanism that we investigate is our proposed independent accumulator (IA) model (see Fig.~\ref{fig:ia-sketch}).
We use the term `independent' to refer to the fact that there are no direct interactions between each accumulator, unlike in the LCA model which has direct competition between states.
To enable a form of competition, we add a second thresholding layer that  projects back to self-excite and mutually-inhibit the first layer.
We now provide the details of each layer~-- again implemented using the principles of the NEF\@.
\begin{figure}
    \centering
    \includegraphics[scale=0.9]{figures/ia-sketch}
    \caption{
        Independent accumulator (IA) network.
        Neural populations are denoted by circles labelled with their represented state variable.
        Arrows denote excitatory connections, while lines ending in circles denote inhibitory connections.
        The second layer computes $\bar{x}_i := \Theta(x_i - \vartheta)$.
    }\label{fig:ia-sketch}
\end{figure}

@@

\section{Results}
We find that the ability to form a clear decision of the LCA network decreases with more noise and less target separation  (see Fig.~\ref{fig:decisions}).
Also, the magnitude of the inputs has an important influence.
For a small inputs with $u=0.2$ the winner will mostly not exceed the 0.15 threshold with noise present.
The best performance is achieved with medium inputs with $u=0.6$.
Higher inputs make it more likely that runner-ups will exceed the 0.15 threshold, especially with a small target separation.
In contrast to the LCA network, the IA network manages to form a clear decision in every trial (not explicitly shown in Figure~\ref{fig:decisions}, but all data points fall on the grey line).
\begin{figure*}[t]
    \centering
    \includegraphics[scale=0.9]{figures/decisions}
    \caption{
        Fraction of trials with a clear decision for the LCA network (which also exactly matches the fraction of correct trials).
        Each plot shows data for a different input magnitude $u$.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal line indicates the optimum, which coincides with the performance of the IA network.
    }\label{fig:decisions}
\end{figure*}

Interestingly, for all clear decisions the correct winner was determined by the LCA network.
Thus, a plot of correct trials looks identical to Figure~\ref{fig:decisions}, with slightly different error bars.
While always reaching a clear decision, the decisions of the IA network are not always correct.
Overall, the IA performance tends to be worse than the LCA performance for a high input magnitude, but better for smaller inputs (Fig.~\ref{fig:correct}a,\,b).
We can greatly improve the IA performance by increasing $\tau_1$ to \SI{0.5}{\second} which slows down the integration of evidence (Fig.~\ref{fig:correct}c).
This improves the IA performance to be above the LCA performance for high baselines, but it will also increase the decisions times of the IA network.
For a low baseline with $u=0.2$, it makes the IA network unable to decide within the allocated time frame, but given more time it would still reach a decision.
%A further improvement to the IA network is possible by eliminating a systematic bias introduced in the second layer of the IA network.
%In the NEF the neuron gain and bias parameters are usually picked from random distributions.
%As such the layer 2 populations will become active at slightly different thresholds.
%By using identical layer 2 populations, i.e., identical gain and bias parameters for each population, this bias can be eliminated and the IA network performs even better (Fig.~\ref{fig:correct}d).
\begin{figure*}[t]
    \centering
    \includegraphics[scale=0.9]{figures/correct}
    \caption{
        Fraction of correct trials for the IA network.
        Each plot shows data for a different combination of input magnitude $u$ and integration time-constant $\tau_1$.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal line indicates the optimum.
    }\label{fig:correct}
\end{figure*}

As shown in Fig.~\ref{fig:time}a, the time required to reach a decision in the LCA network depends mostly on the size of inputs and target separation.
We averaged over the different noise conditions because the noise influence on the timing was minor.
In the IA network the largest input is the most important factor (dashed vs.\ solid lines in Fig.~\ref{fig:time}b).
Depending on this magnitude, the network can either be faster or slower than the LCA network, but it will need more time given a value of $\tau_1$ that achieves the same fraction of correct responses as the LCA network.
There is also a slight influence of target separation and input noise, with an interaction of these two parameters (solid lines in Fig.~\ref{fig:time}).
\begin{figure*}[t]
    \includegraphics[scale=0.9]{figures/time1}
    \hfill
    \includegraphics[scale=0.9]{figures/time2}
    \caption{
        (Left) Mean decision times for the LCA network.
        Shown data is averaged across all noise levels, since noise had minimal effect on decision times.
        (Right) Mean decision times for the IA network with input magnitude $u = 0.2$ (solid) and $u = 1$ (dashed).
        Error bars denote bootstrapped 95\% confidence intervals.
    }\label{fig:time}
\end{figure*}

Finally, looking at the transient responses indicates that both models might produce outputs of losing choices.
For the LCA network the magnitude of the transient response mainly increases with the amount of noise (Fig.~\ref{fig:transient}a).
For the IA network, transient outputs are smaller in noisy conditions, but can be higher than for the LCA network in less noisy conditions with small target separations.
The magnitude of such transient responses is reduced by adjusting $\tau_1$ to 0.5, at the cost of a slower decision.
\begin{figure*}
    \centering
    \includegraphics[scale=0.9]{figures/transient}
    \caption{
        Transient response (highest output of losing choices) for the LCA and IA network.
        Error bars denote bootstrapped 95\% confidence intervals.
        The grey horizontal lines show the optimum.
    }\label{fig:transient}
\end{figure*}

@@

Both models might produce a transient response that may be interpreted as a decision, which can make the detection of decisions problematic.
This is of special relevance when incorporating the networks into larger cognitive models.
For the LCA network, the transient behaviour is inherent to its design; there will always be an initial rise of all state variables (that receive input) before the mutual inhibition grows strong enough to push them back to zero.
In the IA network, however, such transient responses may be avoided by choosing appropriate $\tau_1$ and $\vartheta$.
It should be noted that other recurrently connected readout mechanisms have been proposed for the two choice case that produce bursting outputs somewhat comparable to the IA network~\cite{lo2006}.
However, the evidence integration uses competition by means of pooled inhibition makes it more similar to the LCA than the IA network and the behaviour under highly noisy conditions is unknown.

@@

We have not yet investigated the agreement of these mechanisms with neurobiological and behavioural data, although this has been done before for other WTA networks~\cite{gold2007,smith2004}.
The implementation in spiking neurons, however, provides some basic biological plausibility and more readily permits comparisons with neural data.
In particular, the IA network predicts that the firing rates for neurons in the first layer will rise up to a threshold, and that neurons in the second layer will not become active or inhibit the first layer until reaching this threshold.
Neurons in the macaque lateral intraparietal area give a similar step response~\cite{latimer2015}.
In contrast, the LCA network predicts that the firing of any neurons will proportionately inhibit all other neurons that do not represent the same state.
With regard to behavioural data, different effects for decision times are predicted as the number of choices increases:
for the LCA network decisions become slower due to the mutual interaction, but for the IA network decisions will take less time because only a single accumulator needs to exceed the threshold.
Nevertheless, evidence for one network does not exclude the possibility that the other network is employed for different tasks or by different brain areas.
Relatedly, it might be more plausible to distribute the representation of all state variables over a single population of neurons.
This is directly supported by the NEF, but we chose to leave this to future work to keep the current analysis free from potential interactions of the state variables introduced by such a distributed representation.
